---
output:
  pdf_document: 
      latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 2: Foreign Exchange Rates Confidence Intervals

## Introduction

### Analysis Goal

The goal of this analysis is to explore degrees of uncertainty for Canadian 
foreign exchange rates using confidence intervals. This will be done using 
EUR/CAD and USD/CAD exchange rates. 

Currency exchange rate is important because it determines a nation's economic 
health. In other words, exchange rates informs us about the well-being of a 
country's citizens. The Canadian economic strength affects the value of its 
dollar relative to other countries. Specifically, variables such as interest 
rates, inflation rates, or even demand for financial assets (bonds, stocks, 
CDs).

One might not be interested about exchange rates, or economics in general. 
Nevertheless, any Canadian should know what their Canadian Dollars are worth. 
If you travel in a foreign country, you might need to exchange more or less 
Canadian money to buy foreign currency. If you are interested in diversifying 
your portfolio with foreign assets, you should also be mindful of foreign 
exchange rates, which raises the following question: why is it important for 
Canadian investors to be aware of the statistical notion of Foreign Exchange 
Rates Confidence Intervals. 

We will later extensively define 'confidence intervals' but, in short, it is a 
method which shows a potential range of future prices. To answer the previous 
question, a Canadian investor who is looking to buy foreign assets would benefit 
from knowing the potential range of how much it cost him to buy US dollars if 
he is looking to buy U.S stocks. In this case, the investor would want to know 
the confidence interval for USD/CAD exchange rates, which we will analyze in the 
following sections using a technique called critical values. Additionally, 
confidence intervals are quite relevant during the COVID-19 crisis, as both 
domestic and foreign central banks apply monetary policies, which makes exchange 
rates more volatile. Therefore, we will also need to calculate a confidence 
interval for variance of exchange rates using a technique called bootstrapping 
which will be defined in later sections.

### About Exchange Rates

In order to understand the following analysis, one must understand how exchange 
rates work. The Bank of Canada straightforwardly explains whether a stronger or 
weaker dollar is better. \newline

A stronger dollar:

* is good if your business needs a lot of imports, or you are shopping or 
traveling abroad.
* is bad if your business depends on exports, or on visitors coming to Canada.

A weaker dollar:

* is good if your business depends on exports, or on visitors coming to Canada.
* is bad if your business needs a lot of imports, or if you are shopping or 
traveling abroad. 

> (*1. “Understanding Exchange Rates.”*)  

### Hypothesis


Since quite a large sample will be used for the analysis, we 
hypothesize that the confidence intervals will be close to the truth for both the critical value and empirical bootstrap methods.

## Data

The data that will be used for this analysis are published on Statistics Canada by the Bank of Canada. 
It provides information about 27 different exchange rates for the Canadian 
dollar (*2. Statistics Canada. Daily Average Foreign Exchange Rates*). It has a historical record of daily average prices for all these currencies 
in terms of CAD. The average daily price is recorded because currency prices 
constantly fluctuate.

```{r, echo=FALSE,comment=NA, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(xtable)
library(statcanR)

mydata <- statcan_data("33-10-0036-01", "eng")
```

### Cleaning the data 

The original data had 16 columns with variables such as the reference date, 
scalar id, status, coordinate, geography, currency, or even value. We are only 
interested in the currency, and value variables. The currency column has 27 
different currencies mixed together, which were sorted by date. 

Note the original data has records since 1981 until today. From 1981 to 2017, only the 
"Canadian-Dollar Effective Exchange Rate Index (CERI)" was recorded 
(*3.“Canadian Effective Exchange Rates.”*). It ceased in 2017 which from then on, specific 
currency were recorded from 2017 to today (*4. “BOC Announces Details of Forthcoming Changes”*). Thus we have a sample of specific 
currencies of about 1000 daily averages recorded. There were not more values, 
because weekends are recorded as zeros due to foreign exchange markets being 
closed. We of course removed these values in order to not unnecessarily skew the 
results. There were also about 20 non recorded values which could be due to 
holidays. These were removed. Finally, the data were sorted in order to be left 
with our two currencies of interest for which we created two distinct variables: 
EUR/CAD and USD/CAD. 

```{r, echo=FALSE,comment=NA,message=FALSE}

# EUR/CAD exchange rates

eur <- mydata %>% 
    dplyr::select(Currency = `Type of currency`,
                value = VALUE,
                date = REF_DATE) %>%
  filter(Currency == "European euro, daily average" & value != 0 & !is.na(value))

eurplot <- ggplot(eur, aes(x=value)) + 
  geom_histogram(fill="dodgerblue3", color="white", bins =70) + 
  geom_vline(aes(xintercept=mean(value)),
             color="black", linetype="solid", size=1) +
  labs(x="EUR/CAD",
       y="Count",
       title="European euro, daily averages from 2017-01-03 to Today")
```


```{r, echo=FALSE,comment=NA,message=FALSE}

# USD/CAD exchange rates

usd <- mydata %>% 
    dplyr::select(Currency = `Type of currency`,
                value = VALUE,
                date = REF_DATE) %>%
  filter(Currency == "U.S. dollar, daily average" & value != 0 & !is.na(value))

usdplot<- ggplot(usd, aes(x=value)) + 
  geom_histogram(fill="mediumaquamarine", color="white", bins =70) + 
  geom_vline(aes(xintercept=mean(value)),
             color="black", linetype="solid", size=1) +
  labs(x="USD/CAD",
       y="Count",
       title="U.S. dollar, daily averages from 2017-01-03 to Today")
```

### Definition of the Variables that will be used extensively

Exchange rates are expressed as 1 unit of the foreign currency converted into Canadian dollars.

EUR/CAD: This variable defines how much Canadian dollar is needed in order to 
buy 1 euro on a certain day. For example, as of 2021-03-17, 1.4858 CAD = 1 EUR.

USD/CAD:  This variable defines how much Canadian dollar is needed in order to 
buy 1 U.S dollar on a certain day. For example, as of 2021-03-17, 1.2465 CAD = 1 USD. 

Note that these variables have 4 decimals places as in most cases, the exchange 
rate is quoted to four decimal places (*5. “Background Information on Foreign Exchange Rates.”*). Nevertheless, we will break this rule when analyzing key statistics or confidence intervals for more accurate analysis. 

### Numerical Summary of EUR/CAD and USD/CAD

```{r, echo=FALSE,comment=NA,message=FALSE}

# Numerical summaries with both currencies

options(xtable.comment = FALSE)

temp <- mydata %>% 
    dplyr::select(Currency = `Type of currency`,
                value = VALUE,
                date = REF_DATE) %>%
  filter(Currency =="U.S. dollar, daily average" | Currency =="European euro, daily average") %>%
  filter(value != 0 & !is.na(value))


summaryboth <- xtable(x = temp %>% 
                        group_by(Currency) %>%
                        summarise(
                          Mean = mean(value),
                          Var = var(value),
                          Min = min(value),
                          Q1 = quantile(value, 0.25)[[1]],
                          Median = median(value),
                          Q3 = quantile(value, c(0.75)[[1]]),
                          Max = max(value)),
                      caption = 'Numerical Summaries for EUR/CAD and USD/CAD daily averages',
                      digits = 5
                      )
```

Let's calculate a few statistics about both exchange rates: 

`r summaryboth` 

First, note that the minimum for both EUR/CAD and USD/CAD were recorded 
as 1.3832 EUR and  1.2128 USD for 1 CAD on 2017-02-14 for the euro and 2017-09-11 for the U.S dollar. This is 
approximately during the same period, meaning February 2017 was a time where the Canadian dollar was strong;
Thus Canada was at a higher level of economic health compared to other time intervals.

On the contrary, the max for EUR/CAD was recorded as 1.6124 on 2018-03-19. The 
max for the USD/CAD was recorded as 1.4496 on 2020-03-18. This is interesting 
because there are 2 years between these two dates. We said earlier that monetary 
policy do impact the exchange rates, whether they are applied by the domestic or foreign 
central banks. 

Furthermore, note the means of both currencies. It is 1.5042 for the European euro, daily average 
and, 1.3134 for the  U.S. dollar, daily average. Even though high precision is important for 
currency records, we can still state that both means are close their respective medians of 
1.5072 EUR and 1.3163 USD. These are nice to keep in mind as we will calculate 
confidence intervals for the mean in the next sections. 

Finally, we also take note of the variances for U.S dollar and EUR daily averages. 
They are respectively 0.0019 and 0.0015. The EUR/CAD has higher variance. This is probably due to 
the economic situations of Europe and Canada, varying more relative to each other, and/or their central banks 
applying distinct monetary policies resulting in greater volatility in the EUR/CAD exchange as opposed to the 
USD/CAD. These are also nice to keep in mind as we will calculate confidence intervals for the variance in later sections.

### Visualizing the Data

Let's plot histograms for our data where both currencies have about 1,050 records:

```{r, echo=FALSE,comment=NA,message=FALSE, fig.height= 5.8}
eurplot / usdplot
```

The black vertical lines represent the means calculated earlier. The European euro, 
daily averages graph shows that large chunks of data are spread out between 1.45 to 1.57. 
It is visually hard to tell if the EUR/CAD data are more right or left skewed. The U.S. dollar, 
daily averages graph seems a little right skewed. As opposed to the EUR/CAD, the U.S dollar 
looks more concentrated around its mean. 

The short previous graph analysis did not give us much information. This is why in the next section, 
we will calculate confidence intervals for the mean and variance via critical values and 
empirical bootstrapping respectively. Explanations of all these concepts are in the following 
'Methods' section. 

## Methods: Confidence Intervals 

In one sentence, a confidence interval is a set of values within which we are reasonably confident our true value exists.
More specifically, a confidence interval is how much uncertainty there is with any particular statistic. Confidence intervals are often used with a margin of error. It tells you how confident you can be that the results of a poll or survey reflect what you would expect to find if it were possible to survey the entire population. Confidence intervals are essentially connected to confidence levels.

Mathematically, the notation for a confidence interval is defined as follows: 

> $(l_n, u_n)$ where $l_n = g(x_1,...,x_n)$ and $u_n = h(x_1,...,x_n)$, is called a $100\gamma\%$ 
confidence interval for $\theta$. The number $\gamma$ is called the confidence level.
(*6. Dekking, Michel, et al.*)

This might seem a bit mathematically rigorous, so we will not dive deeper into the theory of confidence intervals, keeping the explanations simple for the general reader to understand. We simply want to find these intervals, but we  still need 
to understand the methods used to do so. Two common methods will be used. 

1. Critical values in order to find a confidence interval for the mean of EUR/CAD and USD/CAD daily averages. The parameter of interest is the mean $\mu$ which is defined by the expected value of a population. The confidence interval will give us a range of where this value will likely fall. 
2. Empirical bootstrapping in order to find a confidence interval for the variance of EUR/CAD and USD/CAD daily averages. The parameter of interest is the variance $\sigma^2$ which is a measure of how far apart a set of numbers are from their mean value. The confidence interval will give us a range of where this value will likely fall. 

### First Method: Critical Values of a confidence interval for the Mean 

There are two ways of calculating the confidence interval $(l_n, u_n)$ for $\mu$ using critical values:
Using the student's distribution $T$ which yields a confidence interval for $\mu$ of $(\bar{x}_n-t\frac{s_n}{\sqrt{n}},\bar{x}_n+t\frac{s_n}{\sqrt{n}})$. This method is often used when the data are not normally distributed, with unknown variance and, the sample size n, is small. Or, using the normal $Z$ distribution which yields a confidence interval for $\mu$ of $(\bar{x}_n-z\frac{\sigma}{\sqrt{n}},\bar{x}_n+z\frac{\sigma}{\sqrt{n}})$. This method is used when variance is known and the sample size n, is very large. 

Our data are not exactly normal, but quite close. We will assume the data are a realization of the normal distribution. 
Additionally, $n$ is very large: $n=1052$. This means that using $Z$ or $T$ will not make a difference due to the central limit theorem. Therefore, we will make the choice of using $Z$.

Finally, our methodology for calculating a confidence interval about the mean using critical values will be as follows:

* Assume the data can be interpreted as the sample realization $X_1,...,X_n$ from a $Normal(\mu,\sigma^2)$ distribution.
* $\mu$ is the unknown parameter of interest.
* The confidence interval $(l_n, u_n)$ will be as follows $(\bar{x}_n-z_{\alpha/2}\frac{s_n}{\sqrt{n}},\bar{x}_n+z_{\alpha/2}\frac{s_n}{\sqrt{n}})$ where $\alpha = 0.01$ in order to calculate a 99% confidence interval for $\mu$. 

These steps will be done to find a 99% confidence interval of the mean for both EUR/CAD and USD/CAD daily averages. 

### Second Method: Empirical Bootstrapping for a confidence interval of Variance

There are different ways to perform bootstrapping. For this analysis, we will use Empirical Bootstrapping. In empirical bootstrapping, given a population, we take a reasonably sized sample, and then we repeatedly take samples of size n from the initial sample and
calculate the statistic of interest for that sample. After calculating a statistic for each sample taken, we find
the interval that contains the middle 90% of these statistics.

In this particular bootstrap, we will find a 90% confidence interval of the variance for both EUR/CAD and USD/CAD daily averages. We will consider the data to be our original sample. Then, from that original data, we take 1000 bootstrap samples of size 1052. This is done with replacement, meaning a value could appear more than once in the bootstrap samples. 

This yields a bootstrap dataset denoted by $$x_1^*,x_2^*,...,x_n^*$$

We now need to calculate the variance of each of these bootstrap samples. This leaves us with a bootstrap sampling distribution of variances. 

Finally, the 90% confidence interval of variances for the currency daily average is defined by the 0.05 quantile and the 0.95 quantile of the bootstrap sampling distribution. Of course, this process is done for both EUR/CAD and USD/CAD. 

## Results

This section applies the two methods we have defined above and seeks to interpret the results. 

### Critical Values for a confidence interval of the Mean

Let's calculate the confidence interval for the population mean using critical values. The results are displayed in the following table:

```{r, echo=FALSE,comment=NA,message=FALSE}
 
# 99% critical values confidence interval for the mean EURO

xbar_eur <- mean(eur$value)
s_eur <- sd(eur$value)
n_eur <- nrow(eur)

ci_eur_low <- xbar_eur - qnorm(0.995)*s_eur/n_eur
ci_eur_up <- xbar_eur + qnorm(0.995)*s_eur/n_eur
```

```{r, echo=FALSE,comment=NA,message=FALSE}
 
# 99% critical values confidence interval for the mean USD


xbar_usd <- mean(usd$value)
s_usd <- sd(usd$value)
n_usd <- nrow(usd) 

ci_usd_low <- xbar_usd - qnorm(0.995)*s_usd/n_usd
ci_usd_up <- xbar_usd + qnorm(0.995)*s_usd/n_usd
```

```{r, echo=FALSE,comment=NA,message=FALSE}

matrix(c(c("European euro, daily averages", "U.S. dollar, daily average"),c(round(ci_eur_low,5), round(ci_usd_low,5)), c(round(xbar_eur,5), round(xbar_usd,5)), c(round(ci_eur_up, 5),round(ci_usd_up,5))),  ncol = 4) %>% knitr::kable(col.names = c("Currency", "CI lower bound", "Sample Mean", "CI upper bound"),caption = '99 % CI for mean using Critical Values', format = "simple")
```

First, lets look at the EUR/CAD 99% confidence interval: (`r round(ci_eur_low,5)`,`r round(ci_eur_up, 5)`). The upper and lower bounds are very close. They can only be differentiated at the fourth decimal place. We have stated multiple times that the sample size is very large which is also the cause for such a small confidence interval with 99% confidence level. 

The 99% confidence interval for USD/CAD is  (`r round(ci_usd_low,5)`,`r round(ci_usd_up, 5)`). Similar to the EUR/CAD, the confidence interval for USD/CAD is very small. Once more, this very small CI is due to the large sample size. 

Note that we have also calculated the sample mean which is not to be confused with the population mean. This is just for reference. Nevertheless, the sample means of both currencies fall within the confidence intervals for the population means. Again, this is due to our sample size being very large which makes it very probable that the sample means are close to the true population means.

### Empirical Bootstraping for a confidence interval of Variance

Let's calculate the confidence interval for the variances using empirical bootstrapping. The results are displayed in the following table:

```{r, echo=FALSE,comment=NA,message=FALSE}

# 90% empirical bootstrap confidence interval variance EUR


boot_vars_eur <- rep(NA, 1000)

for (i in 1:1000){
  boot_samp_eur <- eur %>% sample_n(size = nrow(eur), replace = TRUE)
  boot_vars_eur[i] <- as.numeric(boot_samp_eur %>% summarize(var(value)))
}

eurvarplot <- tibble(Variances=boot_vars_eur) %>% 
  ggplot(aes(x=Variances)) +
  geom_histogram(fill="wheat2", color="black", bins =50)+
  geom_vline(xintercept = quantile(boot_vars_eur, c(0.05, 0.95)), color="red", size = 0.5) +
  labs(title= "Bootstrap Sampling Distribution of Variances for EUR/CAD", x="Variances", y="Count")
```


```{r, echo=FALSE,comment=NA,message=FALSE}

# 90% empirical bootstrap confidence interval variance USD


boot_vars_usd <- rep(NA, 1000)

for (i in 1:1000){
  boot_samp_usd <- usd %>% sample_n(size = nrow(usd), replace = TRUE)
  boot_vars_usd[i] <- as.numeric(boot_samp_usd %>% summarize(var(value)))
}

usdvarplot <- tibble(Variances=boot_vars_usd) %>%
  ggplot(aes(x=Variances)) +
  geom_histogram(fill="wheat3", color="black", bins =50)+
  geom_vline(xintercept = quantile(boot_vars_usd, c(0.05, 0.95)), color="red", size = 0.5) +
  labs(title= "Bootstrap Sampling Distribution of Variances for USD/CAD", x="Variances", y="Count")
```
```{r, echo=FALSE,comment=NA,message=FALSE}
matrix(c(c("European euro, daily averages", "U.S. dollar, daily average"),c(round(quantile(boot_vars_eur,0.05),6), round(quantile(boot_vars_usd,0.05),6)), c(round(var(eur$value),6), round(var(usd$value),6)), c(round(quantile(boot_vars_eur,0.95), 6),round(quantile(boot_vars_usd,0.95),6))),  ncol = 4) %>% knitr::kable(col.names = c("Currency", "CI lower bound", "Sample Variance", "CI upper bound"),caption = '90 % CI for Variance using Empirical Boostrap', format = "simple")

```


With variance as our parameter of interest, the 90% confidence interval for EUR/CAD is (`r round(quantile(boot_vars_eur, c(0.05,0.95)),6)`). This is a reasonable confidence interval which is very likely to be accurate due to the large size of the bootstrap samples. As variance is one way of measuring volatility, this confidence interval is a great suggestion to measure possible fluctuations of the EUR/CAD daily averages given a 90% confidence level (*7. Kuepper, Justin. “Volatility.”*). Note that the sample variance falls in the confidence interval of variance. 

Let's visualize the EUR/CAD bootstrap sampling distribution of variances. The 90% confidence interval is plotted in red. 

```{r, echo=FALSE,comment=NA,message=FALSE}
eurvarplot
```

The bootstrap sampling distribution appears to approach a normal distribution. This is the effect of the central limit theorem. Furthermore, the confidence interval looks wide graphically. This is the result of choosing a high confidence level of 90%. 

Let's now talk about the results for the USD/CAD currency. With variance as our parameter of interest, the 90% confidence interval for USD/CAD daily averages is (`r round(quantile(boot_vars_usd, c(0.05,0.95)),6)`). Again, the confidence interval is very likely to be accurate due to the sample size. Note that the sample variance falls in the confidence interval of variance. We know the sample variance for EUR/CAD prices is higher than USD/CAD. Notice that the confidence interval for EUR/CAD is larger than USD/CAD. This means there is more uncertainty about fluctuation of EUR/CAD prices rather than USD/CAD prices.

Let's visualize the USD/CAD bootstrap sampling distribution of variances. The 90% confidence interval is plotted in red.

```{r, echo=FALSE,comment=NA,message=FALSE}
usdvarplot
```

This bootstrap sampling also appears to be approaching a normal distribution for the same reasons stated earlier. As we have chosen a 90% confidence level, the two red vertical lines seem far apart. Of course, if we had chosen a lower confidence level, these two vertical lines would be closer. 

## Conclusions

In this analysis the variables EUR/CAD and USD/CAD daily averages were analyzed. We have made use of two methods in order to find confidence intervals for different parameters of interest. 

The first parameter of interest was the mean for which we found a confidence interval using critical values. The second parameter of interest was the variance for which we found a confidence interval using bootstrap sampling.

The initial hypothesis was confirmed. As we have used a large sample, the results are close to the truth for both methods. To restate conclusive results found using the bootstrap sampling method, with variance as our parameter of interest, we have found the following confidence intervals:

* EUR/CAD: (`r round(quantile(boot_vars_eur, c(0.05,0.95)),6)`)
* USD/CAD: (`r round(quantile(boot_vars_usd, c(0.05,0.95)),6)`)

These results can be interpreted as measure for possible volatility of the currencies. Nevertheless, there are drawbacks to measuring fluctuations using this single method as there are more tools to measure volatility. An example is standard deviation. Thus, the next step that one could take to further analyze volatility of currency prices is to estimate confidence intervals for the standard deviation. 

\pagebreak

_All analysis for this report was programmed using `R version 4.0.3 (2020-10-10)`._

## Bibliography


1. “Understanding Exchange Rates.” Bank of Canada, [www.bankofcanada.ca/2020/08/understanding-exchange-rates/](www.bankofcanada.ca/2020/08/understanding-exchange-rates/) (Last Accessed: March 17, 2021)

2. Government of Canada, Statistics Canada. Daily Average Foreign Exchange Rates in Canadian Dollars, Bank of Canada, Government of Canada, Statistics Canada, 23 Mar. 2021, [www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3310003601](www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3310003601) (Last Accessed: March 17, 2021)

3. “Canadian Effective Exchange Rates.” Bank of Canada, [www.bankofcanada.ca/rates/exchange/canadian-effective-exchange-rates/](“Canadian Effective Exchange Rates.” Bank of Canada, www.bankofcanada.ca/rates/exchange/canadian-effective-exchange-rates/. ) (Last Accessed: March 17, 2021)

4. “Bank of Canada Announces Details of Forthcoming Changes to Its Published Foreign Exchange Rate Data.” Bank of Canada, [www.bankofcanada.ca/2016/12/bank-canada-announces-details-forthcoming-changes/](www.bankofcanada.ca/2016/12/bank-canada-announces-details-forthcoming-changes/) (Last Accessed: March 18, 2021)

5. “Background Information on Foreign Exchange Rates.” Bank of Canada, [www.bankofcanada.ca/rates/exchange/background-information-on-foreign-exchange-rates/](www.bankofcanada.ca/rates/exchange/background-information-on-foreign-exchange-rates/) (Last Accessed: March 18, 2021)

6. Dekking, Michel, et al. A Modern Introduction to Probability and Statistics: Understanding Why and How. Springer, 2010. (Last Accessed: March 19, 2021)

7. Kuepper, Justin. “Volatility.” Investopedia, Investopedia, 4 Mar. 2021, [www.investopedia.com/terms/v/volatility.asp](www.investopedia.com/terms/v/volatility.asp) (Last Accessed: March 19, 2021)

8. “ggplot2.” Function | R Documentation, [https://www.rdocumentation.org/packages/ggplot2/versions/3.3.3/topics/ggplot](https://www.rdocumentation.org/packages/ggplot2/versions/3.3.3/topics/ggplot). (Last Accessed: March 19, 2021)
